Modelling low Mach number stellar hydrodynamics with MAESTROeX

### 2019-10-01 astronum_2019_MAESTROeX/astronum-maestroex.tex
minor comments/changes

For many flows in astrophysical systems, the magnitude of the fluid velocity is much less than the soundspeed. Consequently, the Mach number, the ratio of the fluid velocity to the sound speed, is much less than 1: $\textrm{Ma} = |\Ub| / c_s \ll 1$. Such low Mach number flows are challenging to model with standard compressible schemes using explicit timestepping methods, where the maximum size of the timestep is determined by the Courant-Friedrich-Lewy (CFL) condition. This states that for a grid with cell spacing $\Delta x$, the timestep $\Delta t <  \Delta x / \max(|\Ub|+c_s)$. For low Mach number flows, this condition is dominated by the contributions of the sound speed, with the result that many fine timesteps are required to achieve high spatial resolution. Long timescale, high resolution simulations therefore become extremely computationally expensive.

@@

The technique that we use in \maestroex~is to modify the fluid equations themselves so as to filter out the soundwaves. This is a similar approach to the incompressible \cite{Boussinesq1901} and anelastic \cite{Ogura1962a,Gough1968,Durran1989} approximations, however the approximation that we use (variously called the low Mach number approximation \cite{Day2000,Almgren2006a,Nonaka2010} and the generalized pseudo-incompressible approximation \cite{Vasil2013}) allows for background stratification and large pressure perturbations due to heating and changes in composition. This is achieved by decomposing the pressure into a 1d hydrostatic base state, $p_0 = p_0(r, t)$, and a dynamic pressure perturbation, $\pi = \pi(r, \bm{x}, t)$, such that the full pressure is given by $p(r, \bm{x}, t) = p_0(r, t) + \pi(r, \bm{x}, t)$. In order for the approximation to hold, it is necessary that $|\pi|/p_0 = O(\mathrm{Ma}^2)$.

@@

The numerical algorithm implemented in \maestroex~improves upon the original \maestro~algorithm in a number of ways. The temporal integration method has been greatly simplified without compromising the second order accuracy, and a new spherical base state mapping has been implemented in order to reduce mapping errors between spherical and Cartesian grids. Utilising the \amrex~framework has allowed us to implement MPI+OpenMP parallelism, which has been shown to scale well to over 10,000 MPI processes.

@@

\maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement hydrodynamics codes for astrophysical flows. Other codes in this family include \castro, an astrophysical radiation hydrodynamics simulation code \cite{Almgren2010}, and \nyx, an N-body hydrodynamics cosmological simulation code \cite{Almgren2013}. All three codes are developed using the \amrex~software framework, and share much in common in terms of their development, structure and numerical methods. Developers of the different codes work closely together (in fact many of the developers work on more than one of the codes), with their shared structure and framework meaning that new features developed in one code can easily be replicated in the others. Examples of this include the SDC solvers \cite{Zingale2019} and GPU port that have recently been implemented in \castro, and are currently in the process of being implemented in \maestroex.

@@

The simplest way to include rotation is to ignore the effects of the centrifugal force, and only include the Coriolis force. This was done in \maestro, and should be a reasonable approximation for slowly rotating systems.

@@

We have begun porting \maestroex~to GPUs, leveraging the GPU capabilities built into \amrex~and following the lessons learnt porting \castro~to GPUs. In the \amrex~framework, routines can be ported to GPUs by simply inserting a few macros into the code. The boilerplate code to required to compile this for GPU is then generated by a custom preprocessor prior to compilation. CUDA managed memory is used to take care of data transfers between the CPU and GPU, so no explicit copies of data to/from the CPU and GPU are required, significantly reducing the extra code required to port the code. One of the key emphases of the \amrex~approach to GPUs is to maintain performance portability: porting code to GPUs should not come at the expense of the code's performance on CPU-only machines. Because GPU-specific code is created by the preprocessor prior to compilation, it is not necessary to maintain separate versions of the code for different architectures. By changing the compiler flags, the same bit of code can be compiled for serial, MPI-only, OpenMP-only, MPI+OpenMP, GPU-only or MPI+GPU.   
In order to achieve the best performance on the GPU, some additional optimization is required. However, it has been found that these optimizations often improve the performance when the code is run on the CPU only as well. 
More details to appear in a forthcoming \amrexastro~GPU paper.

>>>>

For many flows in astrophysical systems, the magnitude of the fluid velocity is much less than the soundspeed. Consequently, the Mach number, the ratio of the fluid velocity to the sound speed, is much less than 1: $\textrm{Ma} = |\Ub| / c_s \ll 1$. Such low Mach number flows are challenging to model with standard compressible schemes using explicit timestepping methods, where the maximum size of the timestep is determined by the Courant-Friedrich-Lewy (CFL) condition. This states that for a grid with cell spacing $\Delta x$, the timestep $\Delta t <  \Delta x / \max(|\Ub|+c_s)$. For low Mach number flows, this condition is dominated by the contributions of the sound speed, with the result that many fine timesteps are required when using high spatial resolution. Long timescale, high resolution simulations therefore become extremely computationally expensive.

@@

The technique that we use in \maestroex~is to modify the fluid equations themselves so as to filter out the soundwaves. This is a similar approach to the incompressible \cite{Boussinesq1901} and anelastic \cite{Ogura1962a,Gough1968,Durran1989} approximations, however the approximation that we use (called the low Mach number approximation \cite{Day2000,Almgren2006a,Nonaka2010} and the generalized pseudo-incompressible approximation \cite{Vasil2013}) allows for background stratification and large density and temperature perturbations due to heating and changes in composition. This is achieved by decomposing the pressure into a 1d hydrostatic base state, $p_0 = p_0(r, t)$, and a dynamic pressure perturbation, $\pi = \pi(r, \bm{x}, t)$, such that the full pressure is given by $p(r, \bm{x}, t) = p_0(r, t) + \pi(\bm{x}, t)$. In the low Mach number regime, asymptotic analysis shows that $|\pi|/p_0 = O(\mathrm{Ma}^2)$.

@@

The numerical algorithm implemented in \maestroex~improves upon the original \maestro~algorithm in a number of ways. The temporal integration method has been greatly simplified without compromising the second order accuracy, and a new spherical base state mapping has been implemented in order to reduce mapping errors between spherical and Cartesian grids. Utilising the \amrex~framework has allowed us to implement MPI+OpenMP (with tiling \cite{Zhang2016}) parallelism, which has been shown to scale well to over 10,000 MPI processes.

@@

\maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement hydrodynamics codes for astrophysical flows. Other codes in this family include \castro, an astrophysical radiation hydrodynamics simulation code \cite{Almgren2010}, and \nyx, an N-body hydrodynamics cosmological simulation code \cite{Almgren2013}. All three codes are developed using the \amrex~software framework, and share much in common in terms of their development, structure and numerical methods. Developers of the different codes work closely together (in fact many of the developers work on more than one of the codes), with their shared structure and framework meaning that new features developed in one code can easily be replicated in the others. A prime example of this is the GPU port capability that has recently been implemented in \castro, and are currently in the process of being implemented in \maestroex. 
\MarginPar{mention starkiller astrophysics}

@@

The simplest way to include rotation is to ignore the effects of the centrifugal force, and only include the Coriolis force. This was done in \maestro, and should be a reasonable approximation for slowly rotating systems. 
\MarginPar{In Zingale 2011, equation (2) has the Coriolis and centrifugal terms.  Include that equation here?}

@@

We have begun porting \maestroex~to GPUs, leveraging the GPU capabilities built into \amrex~and following the lessons learned porting \castro~to GPUs. In the \amrex~framework, routines can be ported to GPUs by simply inserting a few macros into the code. The boilerplate code to required to compile this for GPU is then generated by a custom preprocessor prior to compilation. CUDA managed memory is used to take care of data transfers between the CPU and GPU, so no explicit copies of data to/from the CPU and GPU are required, significantly reducing the extra code required to port the code. One of the key emphases of the \amrex~approach to GPUs is to maintain performance portability: porting code to GPUs should not come at the expense of the code's performance on CPU-only machines. Because GPU-specific code is created by the preprocessor prior to compilation, it is not necessary to maintain separate versions of the code for different architectures. By changing the compiler flags, the same bit of code can be compiled for serial, MPI-only, OpenMP-only, MPI+OpenMP, GPU-only or MPI+GPU.   
In order to achieve the best performance on the GPU, some additional optimization is required. However, it has been found that these optimizations often improve the performance when the code is run on the CPU only as well. 
More details to appear in a forthcoming \amrexastro~GPU paper.
